\Title

### 2013-03-01 synmut.tex
little edits here and there, moved supplement body into separate file such that we can include it into the main text as an appendix for submission

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate of
HIV-1 (HIV in the following)~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
nAb response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
population within a few months (see \figurename~\ref{fig:aftnonsyn}).
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 month earlier, but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected for their effect on the amino acid sequence
of the viral proteins. Conversely, synonymous mutations are commonly used as
approximately neutral markers in studies of viral evolution. Neutral markers are
very useful since their dynamics can be compared to that of putatively
functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes since these processes
often depend on RNA folding. For example, the HIV \rev{} response element (RRE)
in \env{} enhances nuclear export of full length or partially spliced viral
transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse
transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
required for priming reverse transcription (RT) and is bound by a pseudoknotted
RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}.

Even in absence of important RNA structures, synonymous codons do not evolve
completely neutrally. Some codons are favored over others in many species
\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
engineered HIV strains with altered codon usage can in some cases produce more
viral protein, but in general replicate less efficiently
\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
keating_rich_2009}. Codon deoptimization has been suggested as attenuation
strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
Purifying selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
rates of evolution at synonymous sites vary along the HIV genome
\citep{mayrose_towards_2007}. Positive selection through the host adaptive
immune system, however, is restricted to changes in the amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.  We
argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to limited recombination in HIV populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We then compare our observations to computational
models of HIV evolution and derive estimates for the effect of synonymous mutations
 on fitness.  Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular site and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
Strictly speaking, no mutation in the HIV population ever fixes because the 
mutation rate and the population size are large. We define ``fixation'' or ``loss''
by not observing the mutation in the sample.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given that we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
reached a high frequency by chance (genetic drift or hitchhiking), and
we expect that selection drives it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env{} that includes the \shankaregion{} are
deleterious, while outside this region most polymorphisms are roughly
neutral. Note that this does not imply that all synonymous mutations in outside 
\shankaregion{} are neutral -- only those mutations observed at high frequencies, which
have been experiencing selection for some time already, tend to be neutral.

@@

The propensity of nucleotides in the HIV genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used in ref.~\citep{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or where subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

@@

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness cost of some synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles; consistently, HIV does not seem to adapt its codon usage to
its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome. These regions, however, undergo frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
\FIG{aft}, which shows that many mutations change rapidly in frequency as a
flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uniform prior distributions as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

@@

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.  To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV infected
patients, we find $A_\text{syn} \approx -0.2$ for synonymous changes and
$A_\text{nonsyn} \approx 0$ for nonsynonymous changes. In the three simulations
shown in \FIG{simfixpvar}, the fixation probability of synonymous alleles
decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
($A_\text{syn} \sim -0.5$) as their fitness cost increases; the synonymous
diversity plummets as well, as deleterious mutations are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the parameters in
certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous mutations but,
simultaneously, a moderately high synonymous diversity. Specifically,
\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
-0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
high fraction ($\gtrsim 0.8$) of sites has to be deleterious with effect size
$|s_d| \sim 0.002$.  This result fits well the expectation based on the
fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
(i) a substantial depression in $\pfix$ requires pervasive deleterious
mutations, otherwise the majority of observed polymorphisms are neutral and no
depression is observed; (ii) in order to hitchhike, the deleterious effect size
has to be much smaller than the escape rate, otherwise the double mutant has
little or no fitness advantage. Consistent with this argument, larger
deleterious effects in \FIG{simheat} correspond to larger escapes rates. (iii)
mutations with a deleterious effect smaller than approximately $0.001$ behave
neutrally consistent with the typical coalescent times observed in HIV.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\affiliation{\Affiliation}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
nAb response against a particular HIV-1 epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 month earlier, but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence
of viral proteins in a way that reduces antibody binding or epitope presentation. 
Conversely, synonymous mutations are commonly used as
approximately neutral markers in studies of viral evolution. Neutral markers are
very useful since their dynamics can be compared to that of putatively
functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes since these processes
often depend on RNA folding. For example, the HIV-1 \rev{} response element (RRE)
in \env{} enhances nuclear export of full length or partially spliced viral
transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse
transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
required for priming reverse transcription (RT) and is bound by a pseudoknotted
RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}.

Even in absence of important RNA structures, synonymous codons do not evolve
completely neutrally. Some codons are favored over others in many species
\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
engineered HIV-1 strains with altered codon usage can in some cases produce more
viral protein, but in general replicate less efficiently
\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
keating_rich_2009}. Codon deoptimization has been suggested as attenuation
strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
Purifying selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
rates of evolution at synonymous sites vary along the HIV-1 genome
\citep{mayrose_towards_2007}. Positive selection through the host adaptive
immune system, however, is restricted to changes in the amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.  We
argue that synonymous mutations reach high frequencies via genetic hitchhiking
due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We then compare our observations to computational
models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
 on fitness.  Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e., with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular site and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
mutation rate and the population size are large. We define ``fixation'' or ``loss''
by not observing the mutation in the sample.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor at this 
position is currently carrying it, i.e., the mutation frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given that we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
reached a high frequency by chance (genetic drift or hitchhiking), and
we expect that selection drives it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env{} that includes the \shankaregion{} are
deleterious, while outside this region most polymorphisms are roughly
neutral. Note that this does not imply that all synonymous mutations outside 
\shankaregion{} are neutral -- only those mutations observed at high frequencies, which
have experienced selection for some time, tend to be neutral.

@@

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used in ref.~\citep{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or where subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

@@

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness cost of some synonymous mutations, in particular codon
usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome. This region, \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
\FIG{aft}, which shows that many mutations change rapidly in frequency as a
flock.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
alleles at intermediate frequencies fixed or were lost. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects $s_d$ of
the order of $- 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of  synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.002$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

@@

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.  To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
patients, we find $A_\text{syn} \approx -0.2$ for synonymous changes and
$A_\text{nonsyn} \approx 0$ for nonsynonymous changes. In the three simulations
shown in \FIG{simfixpvar}, the fixation probability of synonymous alleles
decreases from the neutral expectation ($A_\text{syn} \approx 0$) to zero
($A_\text{syn} \approx -0.5$) as their fitness cost increases; the synonymous
diversity plummets as well, as deleterious mutations are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the parameters in
certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous mutations but,
simultaneously, a moderately high synonymous diversity. Specifically,
\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
-0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
high fraction ($\gtrsim 0.8$) of sites has to be deleterious with effect size
$|s_d| \sim 0.002$.  This result fits well the expectation based on the
fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
(i) a substantial depression in $\pfix$ requires pervasive deleterious
mutations, otherwise the majority of observed polymorphisms are neutral and no
depression is observed; (ii) in order to hitchhike, the deleterious effect size
has to be much smaller than the escape rate, otherwise the double mutant has
little or no fitness advantage. Consistent with this argument, larger
deleterious effects in \FIG{simsfig} correspond to larger escapes rates. (iii)
mutations with a deleterious effect smaller than approximately $0.001$ behave
neutrally consistent with the typical coalescent times observed in HIV-1.
